Computational methods in landscape ecology

### 2024-09-03 Manuscript.qmd
Move software to corresponding sections

format:
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}

@@

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/03-connectivity.qmd >}}
{{< include Sections/04-sampling.qmd >}}
{{< include Sections/05-simulations.qmd >}}
{{< include Sections/06-spatial-planning.qmd >}}
{{< include Sections/07-genetics.qmd >}}
{{< include Sections/08-software.qmd >}}
{{< include Sections/09-discussion.qmd >}}

@@

\begingroup
\raggedright
::: {#refs}
:::
\endgroup

>>>>

format:
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \usepackage{setspace}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}

@@

{{< include Sections/01-introduction.qmd >}}
{{< include Sections/02-spatial-patterns.qmd >}}
{{< include Sections/03-connectivity.qmd >}}
{{< include Sections/04-sampling.qmd >}}
{{< include Sections/05-simulations.qmd >}}
{{< include Sections/06-spatial-planning.qmd >}}
{{< include Sections/07-genetics.qmd >}}
{{< include Sections/08-discussion.qmd >}}

@@

\footnotesize
\setstretch{1.0}
::: {#refs}
:::

### 2024-09-03 Sections/01-introduction.qmd
Move software to corresponding sections

# Introduction

Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology; @Forman1995LandMosaics] - with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984LandscapeEcology; @Turner1989LandscapeEcology; @Turner2005LandscapeEcology].
Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002KeyIssues; @Wu2013KeyConcepts].

Computational science involves analyzing abstracted core mechanisms of research questions using data and algorithms.
Thus, computational ecology can be defined as computational science that is used to address ecological research questions with focus on complex, adaptive systems and data-driven approaches [@Poisot2019DatabasedSynthesisdriven].
Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in controllable, reproducible, and replicable experimental settings [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Because landscape ecology is a cross-disciplinary field including [e.g., social sciences, geography, or ecology and evolution\; @Wiens1997MetapopulationDynamics; @With2019EssentialsLandscape], the general increase of data availability [@Chi2016BigData; @Jaric2020IEcologyHarnessing; @Wuest2020MacroecologyAge; @Nathan2022BigdataApproaches], or the studied complex landscape systems [@Newman2019ScalingComplexity], there is the need for sophisticated computational methods in order to link heterogeneity patterns and ecological processes.
Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012TenSimple; @Wilson2014BestPractices].

For all computational methods the natural world must be translated into representative data models.
The spatial vector data model is based on points, lines, and polygons features that are defined by their geometry and additional attributes related to them.
Contrastingly, the raster data model is based on evenly-spaced grid cells and discrete or continuous values related to each cell [@Lovelace2019Geocomputation; @With2019EssentialsLandscape].
The choice of the data models, spatial and temporal scales, or the thematic resolution are crucial for the analysis [@Wade2003ComparisonVector; @Lechner2016RecentProgress; @Nedd2021SynthesisLand].

Here, we aim to introduce the latest developments of computational methods in landscape ecology.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive, general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
First, we will introduce recent computational developments of the most important topics of landscape ecology.
This includes, for example, spatial patterns quantification, connectivity and movement, or model simulations.
Second, we will introduce common software tools that implement or are potentially capable of implementing these developments.
However, due to its many advantages [@Powers2019OpenScience], we will focus on open-source software , and specifically scripting languages, i.e., R, Python, and Julia.
Last, we will discuss potential future needs developments of computational methods in landscape ecology.

>>>>

# Introduction {#sec-introduction}

Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology; @Forman1995LandMosaics] - with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984LandscapeEcology; @Turner1989LandscapeEcology; @Turner2005LandscapeEcology].
<!-- MH: This sentence needs some editing to be more concise-->
Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, pattern-process links, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002KeyIssues; @Wu2013KeyConcepts].

Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Prlic2012TenSimple; @Wilson2014BestPractices].
Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
Computational ecology is crucial for the field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @Wiens1997MetapopulationDynamics; @With2019EssentialsLandscape], the general increase of data availability [@Chi2016BigData; @Jaric2020IEcologyHarnessing; @Wuest2020MacroecologyAge; @Nathan2022BigdataApproaches], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
Thus, future progress within the research field will highly depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].

<!-- MH: Given the limited word count I would cut this
For all computational methods the natural world must be translated into representative data models.
The spatial vector data model is based on points, lines, and polygons features that are defined by their geometry and additional attributes related to them.
Contrastingly, the raster data model is based on evenly-spaced grid cells and discrete or continuous values related to each cell [@Lovelace2019Geocomputation; @With2019EssentialsLandscape].
The choice of the data models, spatial and temporal scales, or the thematic resolution are crucial for the analysis [@Wade2003ComparisonVector; @Lechner2016RecentProgress; @Nedd2021SynthesisLand].
-->

Here, we aim to introduce the latest developments of computational methods in landscape ecology.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive, general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
First, we will introduce recent computational developments in the most important topics of landscape ecology.
<!-- This includes, for example, spatial patterns quantification, connectivity and movement, or model simulations. -->
Second, we will introduce common software tools that implement or are potentially capable of implementing these developments.
Lastly, we will discuss potential future directions for the development of computational methods in landscape ecology.

"Open science" involves full transparency and data access at all stages of the research processes and integrative and multi-disciplinary fields, such as landscape ecology, can massively benefit from this concept [@Hampton2015TaoOpen].
Using open source software is one key aspect of open science because it facilitates shareability, reproducibility, and transparency and consequently increases trust in the analyses [@Powers2019OpenScience; @Hesselbarth2021OpensourceTools].
<!-- In this context, script-based workflows have an advantage over workflows involving graphical user interfaces because the former are more shareable and reproducible [@Hampton2015TaoOpen; @Powers2019OpenScience]. -->
Thus, we will focus here on open-source software, specifically scripting languages such as R, Python, and Julia.

### 2024-09-03 Sections/02-spatial-patterns.qmd
Move software to corresponding sections

<!-- jn: should we reference software packages here (or in Software) when they implement the methods we discuss? -->
<!-- MH: I think we decided to not reference software here but in a separate section -->
<!-- MH: Alternatively, if we move the software section to the beginning, we could reference software here -->

@@

Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, they are also critiqued for not being able to realistically represent continuous features of the natural word [@Frazier2017LandscapeMetrics].
Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpen].
Recently also several metrics describing landscape complexity were introduced, e.g., several entropy measures (see following sections).

<!-- https://github.com/r-spatialecology/landcapemetrics -->
<!-- https://fragstats.org/ -->

@@

<!-- jn: should we use equations here? -->
<!-- MH: Personally, I think if possible not -->

<!-- jn: ask T. Stepinski for review of this section later -->

It is mainly derived from information theory and thermodynamics, such as Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi entropy.
Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity) and scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
However, studies show that insights do not only depend on the formulation of the used entropy, but also on the underlining data model, for example the composition of categories or the co-occurrence matrix [@Zhao2019CalculatingSpatial].

@@

<!-- https://github.com/r-spatialecology/landcapemetrics -->

@@

<!-- https://jakubnowosad.com/bespatial/ -->

@@

<!-- Kolmogorov, etc. -->
<!--@Konrad2023; Juhász-Nagy’s information theory-based functions;  https://cran.r-project.org/web/packages/LandComp/index.html -->
<!-- "Fractal dimension seems directly related to entropy." @cushmanThermodynamicsLandscapeEcology2015a -->

@@

Spatial patterns can be compared between different areas, or for the same area at two different moments in time.
Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
This allows to find areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021PatternbasedIdentification].

<!-- https://github.com/Nowosad/motif/ -->
<!-- https://jakubnowosad.com/spquery/ -->
<!-- FOSS4G Europe pap? -->

@@

Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm] and were adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
The gradient surface model can increase the resemblance of the data model and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
<!-- MH: What about Spectral and Wavelet Analysis?

<!-- https://www.digitalsurf.com/free-trial/ -->
<!-- https://github.com/bioXgeo/geodiv -->

@@

<!-- jn: ask T. Matuszek for review of this section later -->

@@

Vector-based metrics also have their limitations.
The most important is computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
<!-- MH: What does topologically correct mean?-->
Another technical issue is the need for the data to be topologically correct, which is often problematic especially for data from different sources.
Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.

<!-- https://github.com/r-spatialecology/vectormetrics -->

>>>>

<!-- JN: should we reference software packages here (or in Software) when they implement the methods we discuss? -->
<!-- MH: I think we decided to not reference software here but in a separate section -->
<!-- MH: Alternatively, if we move the software section to the beginning, we could reference software here -->

@@

Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, they are also critiqued for not being able to realistically represent continuous features of the natural word [@Frazier2017LandscapeMetrics].
Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpen].
Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpen], the *PyLandStats* python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].

@@

<!-- JN: should we use equations here? -->
<!-- MH: Personally, I think if possible not -->
<!-- JN: ask T. Stepinski for review of this section later -->

Entropy measures are mainly derived from information theory and thermodynamics, such as Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi entropy.
Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity) and scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
However, studies show that insights do not only depend on the formulation of the used entropy, but also on the underlining data model, for example the composition of categories or the co-occurrence matrix [@Zhao2019CalculatingSpatial].

@@

@@

@@

Entropy metrics are available in various software such as *landscapemetrics*, *PyLandStats*, *besspatial* [@Nowosad2024BespatialBoltzmann], *rasterdiv* [@Rocchini2021RasterdivInformation], or *LandComp* [@Konrad2023AdaptingMultiscale].

<!-- Kolmogorov, etc. -->
<!-- "Fractal dimension seems directly related to entropy." @cushmanThermodynamicsLandscapeEcology2015a -->

@@

Spatial patterns can be compared between different areas, or for the same area at two different moments in time.
Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
This allows to find areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021PatternbasedIdentification].
R packages for such operations include *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource].
Popular Python libraries for spatial operations include *Geopandas* [@Jordahl2020GeopandasPython], *Shapely*[@Gillies2024ShapelyPython], or *PySal*[@Rey2010PySALPython].
The *GeoStats.jl* [@Hoffimann2018GeoStatsJl] and *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] software provide general spatial data operations in the Julia programming language.

@@

<!-- MH: What about Spectral and Wavelet Analysis? -->

Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm] and were adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
The gradient surface model can increase the resemblance of the data model and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
However, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics.

@@

<!-- JN: ask T. Matuszek for review of this section later -->

@@

Vector-based metrics also have their limitations.
The most important is computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
<!-- MH: What does topologically correct mean?-->
Another technical issue is the need for the data to be topologically correct, which is often problematic especially for data from different sources.
Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.
The *vectormetrics* R package, currently under development [@Matuszek2024VectormetricsLandscape], is specifically designed for vector metrics. 
However, all the general software frameworks mentioned earlier, which support vector data handling, can also be used to calculate basic metrics.

### 2024-09-03 Sections/03-connectivity.qmd
Move software to corresponding sections

<!-- Person responsible: Hesselbarth, ??? -->

@@

Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
The latest developments of resistance surfaces have recently been reviewed [@Dutta2022OverviewComputational].
The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

>>>>

<!-- Person responsible: Hesselbarth, Gerber?, Simpkins? -->

@@

Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
The latest developments of resistance surfaces have recently been reviewed [@Dutta2022OverviewComputational].
The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
Software to model connectivity based on resistance surfaces include *ConScape.jl* [@VanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].

### 2024-09-03 Sections/05-simulations.qmd
Move software to corresponding sections

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024LeastcostNetwork].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@VanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR; @Sciaini2018NLMRLandscapetools], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->

>>>>

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024LeastcostNetwork].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@VanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand; @Cambui2015GradientLandSoftware]
